---
title: "rlang 0.1"
---

```{r setup, include = FALSE}
library("rlang")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

It is with great pleasure that we announce the first release of rlang.
This package provides tools for working with core language features of
R and the tidyverse. You can install it by running:

```{r, eval = FALSE}
install.packages("rlang")
```

(rlang is not currently installed with the tidyverse package, but it will be in the near future.)

rlang includes a large number of tools, and we'll be working to describe and document them clearly in the future. In this blog post, we'll introduce the "tidy evaluation" framework, and discuss some of the design principles that underlie rlang. You can learn more at <http://rlang.tidyverse.org>.

## Tidy evaluation

Tidy evaluation, or __tidyeval__ for short, is a new approach to non-standard
evaluation (NSE) that will be implemented in all tidyverse grammars, including
dplyr, tidyr, and ggplot2. 

Tidyeval is built on top of three key tools:

*   __Quosures__, a data structure that captures both an expression and its
    environment. Quosures are a subtype of formulas that have special
    support in tidyeval grammars. You create quosures with `quo()`, 
    `enquo()` and `quos()`. 

*   __[Quasiquotation][quasiq]__, a tool that lets you "unquote", or 
    evaluate, values in the middle of expressions that are otherwised quoted.

*   Tools for evaluating expressions containing quosures: `eval_tidy()`
    and `as_overscope()`. This is what you will need to create your own
    grammars.
    
The complete system is too much to describe in a blog post, so there are two places to learn more:

* To learn how tidyeval will help you program with data analysis grammars
  read [programming with dplyr][tidyeval-dplyr], a vignette that will be
  included in the upcoming dplyr release.
  
* To learn more about the theory behind tidyeval, read the 
  [tidy evaluation][tidyeval-rlang] vignette included in rlang.

## Features and principles in rlang

Many rlang functions overlap with base R functions: the goal of rlang, like many tidyverse packages, is not to allow you to do fundamentally new things, but to do things with greater ease. One way that rlang makes your life easier is by adopting a consistent set of principles that thread throughout the package.

We describe four important principles below:

* Splicing and unquoting syntax.
* Pattern-matching predicates.
* Naming conventions.
* Comprehensive documentation.

### Splicing and unquoting syntax

All rlang functions taking `...` support a special syntax for splicing
and unquoting. For example, take the `lang()` function which creates 
unevaluated function calls (it's similar to `base::call()`). The first 
argument is the name of the function to call, and the subsequent
arguments are the arguments to that function:

```{r}
lang("foo", x = 1, y = "a", z = TRUE)
```

What happens if you already have the arguments in a list?

```{r}
args <- list(x = 1, y = "a", z = TRUE)
lang("foo", args)
```

You can use the unquote-splice operator, `!!!`, to splice the contents of the list in:

```{r}
lang("foo", splice(args))
```

To use this in your own code, call `dots_list()`:

```{r}
capture_dots <- function(...) {
  dots_list(...)
} 

str(capture_dots(a = 1, b = 2, c = 3))
str(capture_dots(!!! list(a = 1, b = 2), c = 3))
```

Using `dots_list()` means that you don't need to provide an extra argument that takes an explicit list, or relying on your users knowing how to correctly use `do.call()`. 

### Pattern-matching predicates

purrr provides an extensive set of predicate functions like `is_character()` and `is_list()` that make it easy to check that arguments are the type that you expect. 

There are two main differences compared to base R equivalents. Firstly, they are less surprisingly: for example `is_vector(factor("a"))` returns `TRUE` and `is_atomic(NULL)` returns `FALSE`. They also have arguments that allow you to check other properties. For example, you can check that
vectors have a given length:

```{r}
is_list(mtcars)
is_list(mtcars, n = 10)
is_list(mtcars, n = 11)
```

This particularly useful for more complex types like calls where you can check the number of arguments (`n`), the function `name`, or its namespace (`ns`):

```{r}
call <- quote(base::foo(bar, baz))

is_lang(call, n = 3)
is_lang(call, n = 2)

is_lang(call, name = "bar")
is_lang(call, name = "foo")

is_lang(call, ns = "rlang")
is_lang(call, ns = "base")

is_lang(call, "foo", n = 2, ns = "base")
```

### Consistent naming

rlang uses strong naming conventions to make it easier to remember what a function does, to support autocomplete, and to hopefully make it easier to guess the name of a function.

*   Prefixes and suffixes for input and output type:

    rlang tries to follow the general rule that prefixes designate the
    input type of a function while suffixes indicate the output type.
    For instance, `env_bind()` takes an environment while `pkg_env()`
    returns one.

*   Side-effects of setter functions: If an rlang setter starts with `set_`, 
    it means it doesn't have side effects; it returns a modified object. If 
    it starts with `mut_`, it changes its input in place.

*   Constructors. If a constructor takes dots, it is named after the output
    type:

    ```{r}
    env(x = 1)
    chr(x = "a")
    lang("foo", x = NULL)
    ```
    
    On the other hand, if it takes components as formed objects, it is
    prefixed with `new_`:
    
    ```{r}
    new_function(list(x = NULL), quote({ x }))
    ```

*   Scalar versus vectorised functions.

    What's the difference between `has_name()` and `have_name()`? The
    former is a scalar predicate while the latter is vectorised:
    
    ```{r}
    has_name(mtcars, "cyl")
    have_name(mtcars)
    have_name(c(a = 1, 2))
    ```
    
    For that reason, `is_na()` is different from the base R function
    `is.na()`: it is a scalar predicate. On the other hand, `are_na()` is
    a vector predicate.
    
    ```{r}
    x <- c(1L, 2L, NA, 3L)
    is_na(x)
    are_na(x)
    ```
    
    This consistency is a helpful hint to beginners as it's often hard to
    know if a function is vectorised.

### Comprehensive documentation

rlang's documentation is intended to be didactic and introduce
mid-level R programmers to deeper concepts and features of the
language. For instance:

- `?env` provides an introduction to scoping issues in R.

- `?lang` and `?pairlist` explain the structure of R expressions.

- `?cnd_signal`, `?with_handlers`, and `?exiting` go over the
  condition system in R.

Writing good documentation is hard, so expect these to get better over time.

[tidyeval-dplyr]: http://dplyr.tidyverse.org/articles/programming.html
[tidyeval-rlang]: http://rlang.tidyverse.org/articles/tidy-evaluation.html
[quasiq]: http://rlang.tidyverse.org/reference/quasiquotation.html
