```{r, child = "setup.Rmd", include = FALSE}
```

Data-masking is a distinctive feature of R whereby programming is performed directly on a data set, with columns defined as normal objects.

```{r, comment = "#>", collapse = TRUE, error = TRUE}
# Unmasked programming
mean(mtcars$cyl + mtcars$am)

# Referring to columns is an error - Where is the data?
mean(cyl + am)

# Data-masking
with(mtcars, mean(cyl + am))
```

While data-masking makes it easy to program interactively with data frames, it makes it harder to create functions. Passing data-masked arguments to functions requires injection with the [embracing operator][embrace-operator] `{{` or, in more complex cases, the [injection operator][injection-operator] `!!`.


# Why does data-masking require embracing and injection?

Injection (also known as quasiquotation) is a metaprogramming feature that allows you to modify parts of a program. This is needed because under the hood data-masking works by [defusing][topic-defusal] R code to prevent its immediate evaluation.  The defused code is resumed later on in a context where data frame columns are defined.

```{r, comment = "#>", collapse = TRUE, error = TRUE}
my_mean <- function(data, var1, var2) {
  dplyr::summarise(data, mean(var1 + var2))
}

my_mean(mtcars, cyl, am)
```

The problem here is that `summarise()` defuses the R code it was supplied, i.e. `mean(var1 + var2)`.  Instead we want it to see `mean(cyl + am)`. This is why we need metaprogramming, we need to modify that piece of code by injecting the code supplied to the function in place of `var1` and `var2`. The easiest way to achieve that is with `{{`.

```{r, comment = "#>", collapse = TRUE}
my_mean <- function(data, var1, var2) {
  dplyr::summarise(data, mean({{ var1 }} + {{ var2 }}))
}

my_mean(mtcars, cyl, am)
```

Technically, `{{` performs two steps: defusal and injection. It defuses the code supplied to our own function and injects it back into another function. This defuse-and-inject pattern can also be performed in distinct steps with [enquo()] and [`!!`][injection-operator].


# What does "masking" mean?

In normal programming you define objects in the current environment, for instance in the global environment or the environment of the function you are writing.

```{r, comment = "#>", collapse = TRUE}
factor <- 1000

# Can now use `factor` in computations
mean(mtcars$cyl * factor)
```

This environment also contains all functions currently in scope. In a script this includes the functions attached with `library()` calls; in a package, the functions imported from other packages.

If evaluation was performed in a data frame only, we'd lose track of these objects and functions necessary to perform computations.  Hence the data are included in a chain of environments where it comes first and has precedence over the user environment. In other words, it _masks_ the user environment.

The easiest way to see this is to define an object with the same name as a column. The latter has precedence because data-variables come before env-variables:

```{r, comment = "#>", collapse = TRUE}
# Defining an env-variable
cyl <- 1000

# Referring to a data-variable
dplyr::summarise(mtcars, mean(cyl))
```

Note that the tidy eval framework provides [pronouns][.data] to disambiguate between the mask and user contexts. It is often a good idea to use these pronouns in production code. The `.env` pronoun is particularly useful when working with abritrary data frames to prevent a column from inadvertently masking a user variable.

```{r, comment = "#>", collapse = TRUE}
cyl <- 1000

mtcars %>%
  dplyr::summarise(
    mean(.data$cyl),
    mean(.env$cyl)
  )
```

Read more about this in [The data mask ambiguity][topic-data-mask-ambiguity].


# How does data-masking work?

Data-masking relies on three language features:

- [Argument defusing][topic-defusal] (also known as quoting or NSE) with [substitute()] (base R) or [enquo()] and [enquos()] (rlang). R code needs to be defused so it can be evaluated later on in a modified context.

- First class environments. Environments are a special type of list-like object in which defused R code can be evaluated.  The named elements in an environment define objects. Lists and data frames can be transformed to environments:

  ```r
  as.environment(mtcars)
  #> <environment: 0x7febb17e3468>
  ```

- Explicit evaluation with [eval()] (base) or [eval_tidy()] (rlang). When R code is defused, evaluation is interrupted. It can be resumed later on with [eval()]:

  ```{r, comment = "#>", collapse = TRUE}
  expr(1 + 1)

  eval(expr(1 + 1))
  ```

  By default `eval()` and `eval_tidy()` evaluate in the current environment.

  ```{r, comment = "#>", collapse = TRUE, error = TRUE}
  code <- expr(mean(cyl + am))
  eval(code)
  ```

  You can supply an optional list or data frame that will be converted to an environment.

  ```{r, comment = "#>", collapse = TRUE}
  eval(code, mtcars)
  ```

  Evaluation of defused code then occurs in the context of a data mask.


# History

The tidyverse embraced the data-masking approach in packages like ggplot2 and dplyr and eventually developed its own programming framework in the rlang package. None of this would have been possible without the following landmark developments from S and R authors.

- The S language introduced data scopes with [attach()] (Becker, Chambers and Wilks, The New S Language, 1988).

- The S language introduced data-masked formulas in modelling functions (Chambers and Hastie, 1993).

- Peter Dalgaard (R team) wrote the frametools package in 1997. It was later included in R as [base::transform()] and [base::subset()]. This API is an important source of inspiration for the dplyr package. It was also the first apparition of _selections_, a variant of data-masking extended and codified later on in the [tidyselect package](https://tidyselect.r-lib.org/articles/syntax.html).

- In 2000 Luke Tierney (R team) [changed formulas](https://github.com/wch/r-source/commit/a945ac8e) to keep track of their original environments. This change published in R 1.1.0 was a crucial step towards hygienic data masking, i.e. the proper resolution of symbols in their original environments. Quosures were inspired by the environment-tracking mechanism of formulas.

- Luke introduced [base::with()] in 2001.

- In 2006 the [data.table package](https://r-datatable.com) included data-masking and selections in the `i` and `j` arguments of the `[` method of a data frame.

- The [dplyr package](https://dplyr.tidyverse.org/) was published in 2014.

- The rlang package developed tidy eval in 2017 as the data-masking framework of the tidyverse. It introduced the notions of [quosure][topic-quosure], [implicit injection][topic-injection] with `!!`  and `!!!`, and [data pronouns][.data].

- In 2019, injection with `{{` was introduced in [rlang 0.4.0](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/) to simplify the defuse-and-inject pattern. This operator allows beginners to transport data-masked arguments across functions more intuitively.
