```{r, child = "setup.Rmd", include = FALSE}
```

[Data-masking][topic-data-masking] functions require special programming patterns when used inside other functions. In this topic we'll review and compare the different patterns that can be used to solve specific problems.

If you are a beginner, you might want to start with one of these tutorials:

- [Programming with dplyr](https://dplyr.tidyverse.org/articles/programming.html)
- [Using ggplot2 in packages](https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html)


# Choosing a pattern

Two main considerations determine which programming pattern you need to wrap a data-masking function:

1. What behaviour does the _wrapped_ function implement?
2. What behaviour should _your_ function implement?

There are two broad approaches:

- The __forwarding patterns__ with which your function inherits the behaviour of the function it interfaces with.

- The __name patterns__ with which your function takes strings or character vectors of column names.

How you implement forwarding and name patterns depends on the kind of arguments your function takes. Different approaches are needed for single named arguments and multiple arguments in `...`.


# Argument behaviours

In a regular function, arguments can be defined in terms of a _type_ of objects that they accept. An argument might accept a character vector, a data frame, a single logical value, etc. Data-masked arguments are more complex. Not only do they generally accept a specific type of objects (for instance `dplyr::mutate()` accepts vectors), they exhibit special computational behaviours.

- Data-masked expressions (base): E.g. [transform()], [with()].  Expressions may refer to the columns of the supplied data frame.

- Data-masked expressions (tidy eval): E.g. [dplyr::mutate()], [ggplot2::aes()]. Same as base data-masking but with tidy eval features enabled. This includes [injection operators][topic-injection] such as [`{{`][embrace-operator] and [`!!`][injection-operator] and the [`.data`] and [`.env`] pronouns.

- Data masked symbols: Same as data-masked arguments but the supplied expressions must be simple column names. This often simplifies things, for instance this is an easy way of avoiding issues of [double evaluation][howto-double-evaluation].

- [Tidy selections][tidyselect::language]: E.g. [dplyr::select()], [tidyr::pivot_longer()]. This is an alternative to data-masking that supports selection helpers like `starts_with()` or `all_of()`, and implements special behaviour for operators like `c()`, `|` and `&`.

  Unlike data-masking, tidy selection is a domain-specific language. There is in fact no masking at all. Expressions are either evaluated in the data frame (e.g. `c(cyl, am)`), or in the user environment (e.g. `all_of()`, `starts_with()`, and any other expressions). This has implications for inheritance of argument behaviour as we will see below.

- [Dynamic dots][dots_dynamic]: These may be data-masked arguments, tidy selections, or just regular arguments. Dynamic dots support injection of multiple arguments with the [`!!!`][splice-operator] operator as well as name injection with [glue][glue-operator] operators.

To let users know about the capabilities of your function arguments, document them with the following tags, depending on which set of semantics they inherit from:

```
@param foo <[`data-masking`][dplyr::args_data_masking]> What `foo` does.

@param bar <[`tidy-select`][dplyr::args_tidy_select]> What `bar` does.

@param ... <[`dynamic-dots`][rlang::dots_dynamic]> What these dots do.
```


# Forwarding patterns

With the forwarding patterns, the arguments of your function behave the same way as the data-masked arguments they are passed in.


## Embrace with `{{`

The embrace operator [`{{`][embrace-operator] is a forwarding syntax for single arguments. You can forward an argument in data-masked context:

```r
my_summarise <- function(data, var) {
  data %>% dplyr::summarise({{ var }})
}
```

Or in tidyselections:

```r
my_pivot_longer <- function(data, var) {
  data %>% tidyr::pivot_longer(cols = {{ var }})
}
```

The function automatically inherits the behaviour of the surrounding context. For instance arguments forwarded to a data-masked context may refer to columns or use the [`.data`] pronoun:

```r
mtcars %>% my_summarise(mean(cyl))

x <- "cyl"
mtcars %>% my_summarise(mean(.data[[x]]))
```

And arguments forwarded to a tidy selection may use all tidyselect features:

```r
mtcars %>% my_pivot_longer(cyl)
mtcars %>% my_pivot_longer(vs:gear)
mtcars %>% my_pivot_longer(starts_with("c"))

x <- c("cyl", "am")
mtcars %>% my_pivot_longer(all_of(x))
```


## Forward `...`

Simple forwarding of `...` arguments does not require any special syntax since dots are already a forwarding syntax. Just pass them to another function like you normally would. This works with data-masked arguments:

```r
my_group_by <- function(.data, ...) {
  .data %>% dplyr::group_by(...)
}

mtcars %>% my_group_by(cyl = cyl * 100, am)
```

As well as tidy selections:

```r
my_select <- function(.data, ...) {
  .data %>% dplyr::select(...)
}

mtcars %>% my_select(starts_with("c"), vs:carb)
```

Some functions take a tidy selection in a single named argument. In that case, pass the `...` inside `c()`:

```r
my_pivot_longer <- function(.data, ...) {
  .data %>% tidyr::pivot_longer(c(...))
}

mtcars %>% my_pivot_longer(starts_with("c"), vs:carb)
```

Inside a tidy selection, `c()` is not a vector concatenator but a selection combinator. This makes it handy to interface between functions that take `...` and functions that take a single argument.


## Defuse and inject

[`{{`][embrace-operator] and `...` are sufficient for most purposes. Sometimes however, it is necessary to decompose the forwarding action into its two constitutive steps, [defusing][topic-defusal] and [injecting][topic-injection].

`{{` is the combination of [enquo()] and [`!!`][injection-operator]. These functions are completely equivalent:

```r
my_summarise <- function(data, var) {
  data %>% dplyr::summarise({{ var }})
}
my_summarise <- function(data, var) {
  data %>% dplyr::summarise(!!enquo(var))
}
```

Passing `...` is equivalent to the combination of [enquos()] and [`!!!`][splice-operator]:

```r
my_group_by <- function(.data, ...) {
  .data %>% dplyr::group_by(...)
}
my_group_by <- function(.data, ...) {
  .data %>% dplyr::group_by(!!!enquos(...))
}
```

The advantage of decomposing the steps is that you gain access to the [defused expressions][topic-defusal]. Once defused, you can inspect or modify the expressions before injecting them in their target context.


### Inspecting input labels

For instance, here is how to create an automatic name for a defused argument using [as_label()]:

```{r, comment = "#>", collapse = TRUE}
f <- function(var) {
  var <- enquo(var)
  as_label(var)
}

f(cyl)

f(1 + 1)
```

With multiple arguments, use the plural variant [enquos()]. Set `.named` to `TRUE` to automatically call [as_label()] on the inputs for which the user has not provided a name (the same behaviour as in most dplyr verbs):

```{r, comment = "#>", collapse = TRUE}
g <- function(...) {
  vars <- enquos(..., .named = TRUE)
  names(vars)
}

g(cyl, 1 + 1)
```

Just like with `dplyr::mutate()`, the user can override automatic names by supplying an explicit name:

```{r, comment = "#>", collapse = TRUE}
g(foo = cyl, bar = 1 + 1)
```

This technique of names inspection is used in the pattern that bridges data-mask behaviour to a tidyselection context (see below). More complicated defuse-and-inject patterns are explored in the Transforming inputs section.


# Names patterns

The names patterns provide an escape hatch for [defused arguments][topic-defusal] that makes it easy to program with strings and character vectors containing the names of data-variables.

Whereas the forwarding patterns are exclusively used within a function, the names patterns can be used anywhere. In a script, you can loop over a character vector with `for` or `lapply()` and use the [`.data`] pattern to connect a name to its data-variable, or directly supply the whole vector to `all_of()`. In a function, using the names patterns on function arguments lets users supply regular data-variable names without the complications that come with data-masking.


## Subsetting the `.data` pronoun

The [`.data`] pronoun is a tidy eval feature that is enabled in all data-masked arguments, just like [`{{`][embrace-operator]. The pronoun represents the data mask and can be subsetted with `[[` and `$`.

```{r, comment = "#>", collapse = TRUE}
mtcars %>% dplyr::summarise(mean = mean(cyl))

mtcars %>% dplyr::summarise(mean = mean(.data$cyl))

var <- "cyl"
mtcars %>% dplyr::summarise(mean = mean(.data[[var]]))
```

The `.data` pronoun can be subsetted in loops:

```{r, comment = "#>", collapse = TRUE}
vars <- c("cyl", "am")

for (var in vars) print(dplyr::summarise(mtcars, mean = mean(.data[[var]])))

purrr::map(vars, ~ dplyr::summarise(mtcars, mean =  mean(.data[[.x]])))
```

And it can be used to connect function arguments to a data-variable:

```{r, comment = "#>", collapse = TRUE}
my_mean <- function(data, var) {
  data %>% dplyr::summarise(mean = mean(.data[[var]]))
}

my_mean(mtcars, "cyl")
```

With this implementation, `my_mean()` is completely insulated from data-masking behaviour and is called like an ordinary function.

```{r, comment = "#>", collapse = TRUE}
# No masking
am <- "cyl"
my_mean(mtcars, am)

# Programmable
my_mean(mtcars, tolower("CYL"))
```


## Character vector of names

The `.data` pronoun can only be subsetted with single column names. It doesn't support single-bracket indexing:

```{r, comment = "#>", collapse = TRUE, error = TRUE}
mtcars %>% dplyr::summarise(.data[c("cyl", "am")])
```

There is no plural variant of `.data` built in tidy eval. Instead, we'll used the `all_of()` operator available in tidy selections to supply character vectors. This is straightforward in functions that take tidy selections, like `tidyr::pivot_longer()`:

```{r, comment = "#>", collapse = TRUE}
vars <- c("cyl", "am")
mtcars %>% tidyr::pivot_longer(all_of(vars))
```

If the function does not take a tidy selection, it might be possible to use a _bridge pattern_. This option is presented in the bridge section below. If a bridge is impossible or inconvenient, a little metaprogramming with the symbolise-and-inject pattern can help.


## Symbolise and inject

The symbolise-and-inject pattern is a fallback names pattern for cases where `across(all_of())` is not supported. It consists in creating [defused expressions][topic-defusal] that refer to the data-variables represented in the names vector. These are then injected in the data mask context.

You can symbolise a character vector with [syms()] or [data_syms()]. Simple symbols returned by `syms()` work in a wider variety of cases (e.g. with base functions) but we'll use `data_syms()` as it is more robust (see Data mask ambiguity section below). It creates calls that subset the [`.data`] pronoun:

```{r, comment = "#>", collapse = TRUE}
vars <- c("cyl", "am")

syms(vars)

data_syms(vars)
```

A list of symbols can be injected in data-masked dots with [`!!!`][splice-operator], the splice operator. It injects each element of the list as a separate argument:

```r
my_group_by <- function(data, vars) {
  data %>% dplyr::group_by(!!!data_syms(vars))
}

my_group_by(vars)
```

In more complex case, you might want to add R code around the symbols. This requires _transformation_ patterns, see the section below.


# Bridge patterns

Sometimes the function you would like to wrap does not implement the behaviour you would like to give to an argument. To work around this may require a little thought since there is no systematic way of turning one behaviour into another. The general technique consists in forwarding the arguments inside a context that implements the behaviour that you want. Then, find a way to bridge the result to the target verb or function.


## `across()` as a selection to data-mask bridge

dplyr 1.0 added support for tidy selections in all verbs via `across()`. This function is normally used for mapping over columns but can also be used without action to perform a simple selection. For instance, if you'd like to pass an argument to `group_by()` with a tidy-selection interface instead of a data-masked one, use `across()` as a bridge:

```r
my_group_by <- function(data, var) {
  data %>% dplyr::group_by(across({{ var }}))
}

mtcars %>% my_group_by(starts_with("c"))
```

Since `across()` takes selections in a single argument (unlike `select()` which takes multiple arguments), you can't directly pass `...`. Instead, take them within `c()`, which is the tidyselect way of supplying multiple selections within a single argument:

```r
my_group_by <- function(.data, ...) {
  .data %>% dplyr::group_by(across(c(...)}))
}

mtcars %>% my_group_by(starts_with("c"), vs:gear)
```


## `across(all_of())` as a names to data mask bridge

If instead of forwarding variables in `across()` you pass them to `all_of()`, you create a names to data mask bridge.

```r
my_group_by <- function(data, vars) {
  data %>% dplyr::group_by(across(all_of(vars)))
}

mtcars %>% my_group_by(c("cyl", "am"))
```

Use this bridge technique to connect vectors of names to a data-masked context.


## `mutate()` as a data-mask to selection bridge

Passing data-masked arguments to a tidy selection is a little more tricky and requires a three step process that involves the defuse-and-inject pattern. 

```r
my_pivot_longer <- function(data, ...) {
  # Defuse `...` and inspect the supplied names
  dots <- enquos(..., .named = TRUE)
  names <- names(dots)

  # Inject in `mutate()` to update the data
  data <- data %>% dplyr::mutate(!!!dots)

  # Select `...` inputs by name with `all_of()`
  data %>%
    tidyr::pivot_longer(cols = all_of(names))
}

mtcars %>% my_pivot_longer(cyl, am = am * 100)
```

1. In a first step we defuse the `...` expressions so that we can inspect their names. The `.named` argument of [enquos()] causes the defused `...` expressions to be automatically named, just like they would be if passed directly to `mutate()`.

2. Once we have the names, we inject the arguments into `mutate()` to process the data-masked arguments and update the data frame.

3. Finally, we pass the names to the tidy selection via [`all_of()`](https://tidyselect.r-lib.org/reference/all_of.html).


# Transforming inputs

## Named inputs versus `...`

In the case of a named argument, we can transform the embraced input simply by surrounding it in more R code. For instance, the `my_summarise()` function is not exactly useful compared to just calling `summarise()`:

```r
my_summarise <- function(data, var) {
  data %>% dplyr::summarise({{ var }})
}
```

We can make it more useful by adding code around the variable:

```r
my_mean <- function(data, var) {
  data %>% dplyr::summarise(mean = mean({{ var }}, na.rm = TRUE))
}
```

For inputs in `...` however, this technique does not work. We would need some kind of templating syntax for dots that lets us specify R code with a placeholder for the dots elements. This isn't built in tidy eval but you can use operators like `dplyr::across()`, `dplyr::if_all()`, or `dplyr::if_any()`. When that isn't possible, you can template the expression manually.


## Transforming inputs with `across()`

The `across()` operation in dplyr is a convenient way of mapping an expression across a set of inputs. We will create a variant of `my_mean()` that computes the `mean()` of all arguments supplied in `...`. The easiest way it to forward the dots to `across()` (which causes `...` to inherit its tidy selection behaviour):

```{r, comment = "#>", collapse = TRUE}
my_mean <- function(data, ...) {
  data %>% dplyr::summarise(across(c(...), ~ mean(.x, na.rm = TRUE)))
}

mtcars %>% my_mean(cyl, carb)

mtcars %>% my_mean(foo = cyl, bar = carb)

mtcars %>% my_mean(starts_with("c"), mpg:disp)
```


## Transforming inputs with `if_all()` and `if_any()`

`dplyr::filter()` requires a different operation than `across()` because it needs to combine the logical expressions with `&` or `|`. To solve this problem dplyr introduced the `if_all()` and `if_any()` variants of `across()`.

In the following example, we filter all rows for which a set of variables are not equal to their minimum value:

```r
filter_non_baseline <- function(.data, ...) {
  .data %>% dplyr::filter(if_all(c(...), ~ .x != min(.x, na.rm = TRUE)))
}

mtcars %>% filter_non_baseline(vs, am, gear)
```


## Transforming inputs manually

If `across()` and variants are not available, you will need to transform the inputs yourself using metaprogramming techniques. A good place to start if you're new to programming on the language is the [Metaprogramming chapter](https://adv-r.hadley.nz/metaprogramming.html) of the advanced R book.

To illustrate the technique we'll reimplement `my_mean()` and `filter_non_baseline()` without using `across()` or `if_all()`. The pattern consists in defusing the input expression, building larger calls around them, and finally inject the modified expressions inside the data-masked functions. We'll start with a single named argument for simplicity:

```{r, comment = "#>", collapse = TRUE}
my_mean <- function(data, var) {
  # Defuse the expression
  var <- enquo(var)

  # Wrap it in a call to `mean()`
  var <- expr(mean(!!var, na.rm = TRUE))

  # Inject the expanded expression
  data %>% dplyr::summarise(mean = !!var)
}

mtcars %>% my_mean(cyl)
```

The `...` technique is similar but a little more involved. We'll use the plural variants `enquos()` and `!!!`. We'll also loop over the variable number of inputs using `purrr::map()`. But the pattern is basically the same:

```{r, comment = "#>", collapse = TRUE}
my_mean <- function(.data, ...) {
  # Defuse the dots. Make sure they are automatically named.
  vars <- enquos(..., .named = TRUE)

  # Map over each defused expression and wrap it in a call to `mean()`
  vars <- purrr::map(vars, ~ expr(mean(!!.x, na.rm = TRUE)))

  # Inject the expressions
  .data %>% dplyr::summarise(!!!vars)
}

mtcars %>% my_mean(cyl)
```

Note that we are inheriting the data-masked behaviour of `summarise()` because we have effectively forwarded `...` inside that verb. This is different than the previous implementation where we inherited the tidy selection behaviour of `across()`. See how we can create vectors on the fly for instance:

```{r, comment = "#>", collapse = TRUE}
mtcars %>% my_mean(cyl = cyl * 100)
```
