```{r, child = "setup.Rmd", include = FALSE}
```

Starting with rlang 1.0, `abort()` includes the erroring function in the message by default:

```{r, error = TRUE}
my_function <- function() {
  abort("Can't do that.")
}

my_function()
```

This works well when `abort()` is called directly within the failing function. However, when the `abort()` call is exported to another function (which we call an "error helper"), we need to be explicit about which function `abort()` is throwing an error for.


# Passing the user context

There are two main kinds of error helpers:

- Simple `abort()` wrappers. These often aim at adding classes and attributes to an error condition in a structured way:

  ```{r}
  stop_my_class <- function(message) {
    abort(message, class = "my_class")
  }
  ```

- Input checking functions. An input checker is typically passed an input and an argument name. It throws an error if the input doesn't conform to expectations:

  ```{r}
  check_string <- function(x, arg = "x") {
    if (!is_string(x)) {
      cli::cli_abort("{.arg {arg}} must be a string.")
    }
  }
  ```

In both cases, the default error call is not very helpful to the end user because it reflects an internal function rather than a user function:

```{r}
my_function <- function(x) {
  check_string(x)
  stop_my_class("Unimplemented")
}
```

```{r, error = TRUE}
my_function(NA)
```

```{r, error = TRUE}
my_function("foo")
```

To fix this, let `abort()` know about the function that it is throwing the error for by passing the corresponding function environment as the `call` argument:

```{r}
stop_my_class <- function(message, call = caller_env()) {
  abort(message, class = "my_class", call = call)
}

check_string <- function(x, arg = "x", call = caller_env()) {
  if (!is_string(x)) {
    cli::cli_abort("{.arg {arg}} must be a string.", call = call)
  }
}
```

```{r, error = TRUE}
my_function(NA)
```

```{r, error = TRUE}
my_function("foo")
```


## Input checkers and `caller_arg()`

The `caller_arg()` helper is useful in input checkers which check an input on the behalf of another function. Instead of hard-coding `arg = "x"`, and forcing the callers to supply it if `"x"` is not the name of the argument being checked, use `caller_arg()`.

```{r}
check_string <- function(x,
                         arg = caller_arg(x),
                         call = caller_env()) {
  if (!is_string(x)) {
    cli::cli_abort("{.arg {arg}} must be a string.", call = call)
  }
}
```

It is a combination of `substitute()` and `rlang::as_label()` which provides a more generally applicable default:

```{r, error = TRUE}
my_function <- function(my_arg) {
  check_string(my_arg)
}

my_function(NA)
```


# Side benefit: backtrace trimming

Another benefit of passing `caller_env()` as `call` is that it allows `abort()` to automatically hide the error helpers

```{r}
my_function <- function() {
  their_function()
}
their_function <- function() {
  error_helper1()
}

error_helper1 <- function(call = caller_env()) {
  error_helper2(call = call)
}
error_helper2 <- function(call = caller_env()) {
  if (use_call) {
    abort("Can't do this", call = call)
  } else {
    abort("Can't do this")
  }
}
```

```{r, error = TRUE}
use_call <- FALSE
their_function()
```

```{r}
rlang::last_error()
```

With the correct `call`, the backtrace is much simpler and lets the user focus on the part of the stack that is relevant to them:

```{r, error = TRUE}
use_call <- TRUE
their_function()
```

```{r}
rlang::last_error()
```


# testthat workflow

Error snapshots are the main way of checking that the correct error call is included in an error message. However you'll need to opt into a new testthat display for warning and error snapshots. With the new display, these are printed by rlang, including the `call` field. This makes it easy to monitor the full appearance of warning and error messages as they are displayed to users.

This display is not applied to all packages yet. With testthat 3.1.2, depend explicitly on rlang >= 1.0.0 to opt in. Starting from testthat 3.1.3, depending on rlang, no matter the version, is sufficient to opt in. In the future, the new display will be enabled for all packages.

Once enabled, create error snapshots with:

```{r, eval = FALSE}
expect_snapshot(error = TRUE, {
  my_function()
})

expect_snapshot_error(my_function())
```

You'll have to make sure that the snapshot coverage for error messages is sufficient for your package.
