---
title: "Chained promises"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chained promises}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Take the following example:

```{r}
library(lazyeval)
f1 <- function(x) lazy(x)
g1 <- function(y) f1(y)

g1(a + b)
```

`lazy()` returns `a + b` because it always tries to find the top-level promise.

In this case the process looks like this:

1. Find the object that `x` is bound to.
2. It's a promise, so find the expr it's bound to (`y`, a symbol) and the 
   environment in which it should be evaluated (the environment of `g()`).
3. Since `x` is bound to a symbol, look up its value: it's bound to a promise.
4. That promise has expression `a + b` and should be evaluated in the global 
   environment.
5. The expression is not a symbol, so stop.

Occasionally, you want to avoid this recursive behaviour, so you can use `follow_symbol = FALSE`:

```{r}
f2 <- function(x) lazy(x, .follow_symbols = FALSE)
g2 <- function(y) f2(y)

g2(a + b)
```

Either way, if you evaluate the lazy expression you'll get the same result:

```{r}
a <- 10
b <- 1

lazy_eval(g1(a + b))
lazy_eval(g2(a + b))
```
