<!-- 
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Lazy: a new approach to NSE}
-->

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Lazy: A new approach to non-standard evalution

This document outlines a new approach to non-standard evaluation. There are three key ideas:

* Instead of using `substitute()`, use `lazy()` to capture both expression
  and environment.
  
* Every function that uses NSE should have a standard evaluation partner
  that does all the work. This function always ends with `_`.
  
* The SE-partner has a flexible input specification to make it easy for people
  to program with.

## `lazy()`

The key tool that makes this approach possible is `lazy()`, an equivalent to `substitute()` that captures both expression and environment of the promise associated with a function argument:

```{r}
library(lazy)
f <- function(x = a - b) {
  lazy(x)
}
f()
f(a + b)
```

As a complement to `eval()`, the lazy package provides `lazy_eval()` that uses the environment associated with the lazy object:

```{r}
a <- 10
b <- 1
lazy_eval(f())
lazy_eval(f(a + b))
```

The second argument to lazy eval is a list or data frame where names should be looked up first:

```{r}
lazy_eval(f(), list(a = 1))
```

`lazy_eval()` also works with formulas, since they contain the same information as a lazy object: an expression (only the RHS is used by convention) and an environment:

```{r}
h <- function(i) {
  ~ 10 + i
}
lazy_eval(h(1))
```

## Standard evaluation

Whenever we need a function that does non-standard evaluation, always write the standard evaluation version first. For example, let's implement our own version of `subset()`:

```{r}
subset2_ <- function(df, condition) {
  r <- lazy_eval(condition, df)
  r <- r & !is.na(r)
  df[r, , drop = FALSE]
} 

subset2_(mtcars, lazy(mpg > 31))
subset2_(mtcars, ~mpg > 31)
```

To make the NSE version as flexible as possible, use `as.lazy()` to coerce input into a lazy object. In general, this requires an environment, and the `parent.frame()` is a reasonable guess. This allows the user to pass in quoted objects and strings. This is a little risky but provides useful scaffolding when learning how to do NSE the right way.

```{r}
subset2_ <- function(df, condition) {
  condition <- as.lazy(condition, parent.frame())
  r <- lazy_eval(condition, df)
  r <- r & !is.na(r)
  df[r, , drop = FALSE]
} 

subset2_(mtcars, quote(mpg > 31))
subset2_(mtcars, "mpg > 31")
```

## Non-standard evaluation

With the standard evaluation version in hand, writing the standard evaluation version is easy. We just use `lazy()` to capture the unevaluated expression and corresponding environment:

```{r}
subset2 <- function(df, condition) {
  condition <- lazy(condition)
  subset2_(df, condition)
}
subset2(mtcars, mpg > 31)
```
